---
title: " "
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 1
    #code_folding: hide
---
###Correlation

Correlaton refers to the strength of the relationship between two different variables. 

Correlation is often represented by the letter r. 

The equation for correlation(r) looks a little intimidating but if it is broken down into steps, it is much more managable. 

$r = \frac{n(\sum(xy)-(\sum x)(\sum(y))}{\sqrt{n \sum x ^2-(\sum x)^2][n \sum y ^2-(\sum y)^2]}}$

Let's take a closer look at the denominator first:

1. $n(\sum(xy)-(\sum x)(\sum(y))$

The first thing we should do is have a dataset to work with:

x = 5, 5, 4, 4, 5, 5, 4, 5, 3, 5, 6, 6, 5, 6, 6
y = 13, 12, 13, 14, 13, 12, 11, 12, 11, 12, 13, 12, 12, 13, 13

Each dataset has 15 data points. When you work with larger datasets you may want to avoid having to count each number so r has a built in function called `length()` which will return the number of items in your variable. 
We know our first term is 15 so we will add that to the formula.

2. $r = \frac{15(\sum(xy)-(\sum x)(\sum(y))}{\sqrt{15 \sum x ^2-(\sum x)^2][15 \sum y ^2-(\sum y)^2]}}$

The next part is asking us to compute the sum of the values of x multiplied by the values of y and then add that to the formula. 

```{r}
x = c(5, 5, 4, 4, 5, 5, 4, 5, 3, 5, 6, 6, 5, 6, 6)
y = c(13, 12, 13, 14, 13, 12, 11, 12, 11, 12, 13, 12, 12, 13, 13)
xy=sum(x*y)
xy
```
3. $r = \frac{15(921)-(\sum x)(\sum(y))}{\sqrt{15 \sum x ^2-(\sum x)^2][15 \sum y ^2-(\sum y)^2]}}$

Next we need to subtract this amount from the result from the product of the sum of all of the values of x and the sum of all the values of y. 

```{r}
x = c(5, 5, 4, 4, 5, 5, 4, 5, 3, 5, 6, 6, 5, 6, 6)
y = c(13, 12, 13, 14, 13, 12, 11, 12, 11, 12, 13, 12, 12, 13, 13)
sumx = sum(x)
sumy = sum(y)
sumx
sumy
```
4. $r = \frac{15(921-(74)(186)}{\sqrt{15 \sum x ^2-(\sum x)^2][15 \sum y ^2-(\sum y)^2]}}$

Let's just take care of the numerator for now.

```{r}
num=(15*xy)-(sumx*sumy)
num
```
  
5. $r = \frac{51}{\sqrt{15 \sum x ^2-(\sum x)^2][15 \sum y ^2-(\sum y)^2]}}$

Alright, now let us look at the denominator. 

${\sqrt{15 \sum x ^2-(\sum x)^2][15 \sum y ^2-(\sum y)^2]}}$

You should notice that the denominator is sort of a mirror image, the first part handles x and the second part handles y. 

```{r}
dsumx=sum(x^2)
dsumy=sum(y^2)
dsumx
dsumy
```
6. $r = {\sqrt{15(376)-(\sum x)^2][15(2316))-(\sum y)^2]}}$

Again, the same thing is being done to x and y. We are going to sum the values in x and then square them, and do the same to the values in y. 

```{r}
ssx=sum(x)^2
ssy=sum(y)^2
ssx
ssy
```
7. ${\sqrt{15(376)-5,476][15(2316))-34,596]}}$

Now we can clean up the denominator:

```{r}
den = sqrt(((15*dsumx)-(ssx))*((15*dsumy)-(ssy)))
den
```
8. $r = \frac{51}{153.675}$

Now divide

```{r}
r = num/den
r
```

9. $r = 0.33$

Before we start discussing what that means, let us see the other way of calculating r (there are many)

```{r}
x
y
cor(x,y)
```

That seems *a lot* easier than the eight steps we went through!

The important thing to remember is that while R can do things for you very fast, it is important that you know why you are doing them in the first place!

So we found that for our data, the r was equal to `r`. This is a relatively weak correlation. In psychology you will likely never see correlations that are above .9. This obtained r would indicate some positive correlation.  

Here is what a plot of the data looks like:

```{r}
plot(x~y)
```
It would be hard to guess a correlation by just looking at this data. There are only 15 points, but we can tell they are not all in the same space, nor are they all randomly dispersed. So if we *had* to take a guess, we could assume that a slight correlation could be prsent. 

Let us take a look at a data set with some meaning and more data points. 

```{r}
plot(cars$speed~cars$dist, xlab="Distance",ylab = "Speed",main="Distance vs. Speed", sub = "r = .81")
cor(cars$speed,cars$dist)
```